Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-08-10 chapters/Paper_two.tex
All references polished, conclusion missing

Wetting phenomena have an impact on areas as diverse as surface coatings, printing, plant treatment with pesticides, or even pandemics~\cite{oronLongscaleEvolutionThin1997, doi:10.1146/annurev-fluid-011212-140734, cassieWettabilityPorousSurfaces1944, Lenormand_1990,DERYCK1998278, doi:10.1146/annurev.fluid.31.1.347,BergeronNature,BhardwajPoF2020}.

@@

If this affinity is not sufficiently strong or the substrate is rather heterogeneous the liquid layer may eventually become unstable,
leading to film rupture~\cite{oronLongscaleEvolutionThin1997, crasterDynamicsStabilityThin2009} or wrinkling~\cite{dasilvasobrinhoStudyDefectsUltrathin1999}. 

@@

A successful mathematical modelling of the dynamics of thin liquid films is based on the thin-film equation~\cite{Reynolds, oronLongscaleEvolutionThin1997}

@@

The function $M(h)$ denotes the mobility, whose
functional form is determined by the fluid velocity boundary condition at 
the solid surface: for a no-slip condition~\cite{oronLongscaleEvolutionThin1997}, in particular, $M(h)=\frac{h^3}{3\mu}$, where $\mu$ is the fluid dynamic viscosity.

@@

where $\gamma$ is the surface tension. The disjoining pressure
$\Pi(h)$ is the derivative, with respect to the height $h$, of the interfacial potential,
which incorporates the interactions between liquid and substrate (i.e. wetting properties).
The advantage of such approach in comparison with molecule-resolved methods such as Molecular Dynamics~\cite{haile1992molecular, zhangMolecularSimulationThin2019, doi:10.1063/1.1290698, grabow1988thin}
and Density Functional Theory~\cite{PhysRevA.40.2567, tarazona1984simple, meister1985density, hughes2014introduction} is that these 
may quickly become computationally prohibitive, as the size of the film is increased above few nanometres in thickness and to the micrometric scale in the horizontal extension.

@@

For instance, in the context of dewetting, rupture times measured in experiments are shorter than predicted by simulations of the thin-film equation~\cite{herminghaus1996dewetting,herminghaus1998dewetting,beckerComplexDewettingScenarios2003}.

@@

Almost half a century later, a stochastic version of Eq.~(\ref{eq:thin_film}) was derived~\cite{grunThinFilmFlowInfluenced2006, meckeThermalFluctuationsThin2005, davidovitchSpreadingViscousFluid2005}, through a lubrication approximation of the Navier-Stokes-Landau-Lifshitz 
equations of fluctuating hydrodynamics~\cite{landauFluidMechanicsLandau2013}.

@@

The influence of thermal fluctuations on the short and long time morphology of a film has been studied with simulations of the one and two dimensional stochastic thin-film equations~\cite{nesicFullyNonlinearDynamics2015, alizadeh_pahlavan_cueto-felgueroso_hosoi_mckinley_juanes_2018}. 
Recently, new theoretical insights have been gained explaining film rupture as a consequence of the combined action of thermal fluctuations and drainage, both numerically~\cite{shahThermalFluctuationsCapillary2019} and experimentally~\cite{PhysRevLett.125.158001}.

@@

\section{Linear Stability Analysis of the Stochastic Thin-Film Equation}\label{sec:stoch_thin_film}
The one-dimensional stochastic thin-film equation reads~\cite{grunThinFilmFlowInfluenced2006, meckeThermalFluctuationsThin2005}:
\begin{equation}\label{eq:stoch_thin_film_simple}
    \partial_t h = \partial_x \left[\frac{h^3}{3\mu}\partial_x p + \sqrt{\frac{2k_BTh^3}{3\mu}}\mathcal{N}\right],
\end{equation}
where $k_B$ is Boltzmann's constant, $T$ is the temperature and $\mathcal{N}$ is a Gaussian white noise with
\begin{align}\label{eq:correlation}
    \langle\mathcal{N}(x,t)\rangle =&~0,\\ 
    \quad \langle\mathcal{N}(x,t)\mathcal{N}(x',t')\rangle =&~\delta(x-x')\delta(t-t').
\end{align}
In order to understand the influence of thermal fluctuations on the temporal evolution of the film height, we cursory recall the linear stability analysis of Eq.~(\ref{eq:stoch_thin_film_simple}) 
\cite{zhangMolecularSimulationThin2019, diezMetallicthinfilmInstabilitySpatially2016}.
Introducing the deviation $\delta h$ from the mean height $h_0$, such that $h = h_0 + \delta h$ and $\delta h \ll h_0$, 
the linearised stochastic thin-film equation is obtained
\begin{equation}\label{eq:linearstability_realspace}
    \partial_t \delta h = \frac{h_0^3}{3\mu}\Big(\Pi^{\prime}(h_0)\partial_x^2\delta h - \gamma \partial_x^4\delta h\Big) + \sqrt{\frac{2k_BT h_0^3}{3\mu}} \partial_x \mathcal{N},
\end{equation}
where $\Pi^{\prime}(h_0) \equiv \frac{\partial \Pi(h)}{\partial h}\bigg\rvert_{h=h_0}$.
Here, it is assumed that also the noise is small, in the sense that $k_B T \ll \gamma h_0^2$.
To derive a dispersion relation for this system, first the Fourier transforms of the perturbation $\delta h$  
\begin{equation}\label{eq:fourier_delta_h}
    \tilde{\delta h}(q,t) = \int_{-\infty}^{\infty} \delta h(x,t) e^{-ixq} dx,
\end{equation}
and the noise term $\mathcal{N}$
\begin{equation}\label{eq:fourier_noise}
    %\tilde{\mathcal{N}}(\mathbf{q},t) = \int_{-\infty}^{\infty} \mathcal{N}(\mathbf{x},t) e^{-i\mathbf{x}\cdot\mathbf{q}} \dd[2]{\mathbf{x}}.
    \tilde{\mathcal{N}}(q,t) = \int_{-\infty}^{\infty} \mathcal{N}(x,t) e^{-ixq} dx.
\end{equation}
must be performed.

Inserting $\tilde{\delta h}$ and $\tilde{\mathcal{N}}$ in~(\ref{eq:linearstability_realspace}) yields
\begin{equation}\label{eq:stability_fourier}
    \partial_t \tilde{\delta h} = \omega(q)\tilde{\delta h} +i\sqrt{\frac{2k_BTh_0^3}{3\mu}}q\tilde{\mathcal{N}},
\end{equation}
with dispersion relation $\omega(q)$ given by
\begin{equation}\label{eq:dispersion}
    \omega(q) = \frac{1}{t_0}\left[2\left(\frac{q}{q_0}\right)^2 - \left(\frac{q}{q_0}\right)^4\right].
\end{equation}
The maximum, $q_0$, of $\omega(q)$ and the characteristic time, $t_0$, are specific to the substrate the film is deposited on and their expressions are~\cite{fetzerThermalNoiseInfluences2007}
\begin{equation}\label{eq:q0}
    q_0^2 = \frac{1}{2\gamma}\frac{\partial \Pi(h)}{\partial h}\bigg\rvert_{h=h_0},
\end{equation}
and
\begin{equation}\label{eq:t0}
    t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}.
\end{equation}
Solving Eq.~(\ref{eq:stability_fourier}) leads to the following expression for the spectrum~\cite{zhangMolecularSimulationThin2019,meckeThermalFluctuationsThin2005},
$S(q,t) = \langle |\tilde{\delta h}(q,t)|^2\rangle$ (where the brackets stand for the average over the noise):
\begin{equation}\label{eq:structure_factor}
    S(q,t) = S_0(q)e^{2\omega(q)t} + \frac{k_BTh_0^3}{3\mu}\frac{q^2}{\omega(q)}(e^{2\omega(q)t} - 1).
\end{equation}
For $q>\sqrt{2}q_0$, $S(q,t)$ tends to the capillary wave spectrum, 
$S \sim S_{\text{cw}} \propto \frac{k_BT}{\gamma q^2}$, as $t \rightarrow \infty$ (whereas it would decay exponentially to zero in the deterministic case) ~\cite{fetzerThermalNoiseInfluences2007,meckeThermalFluctuationsThin2005}.
Moreover, while the maximum $q_0$ is independent of time in the deterministic case, the maximum $q_m$ of 
(\ref{eq:structure_factor}) approaches $q_0$ from the right as the time increases.
It follows that the most unstable wavelength, $\lambda_{\text{max}} = 2\pi/q_m$, grows with time.
We will come back to this fact in Sec.~\ref{sec:results_two} and show that indeed this behavior is reproduced and observed in our simulations.

\section{Thin Film Lattice Boltzmann Model}\label{sec:num_method}
Performing numerical simulations of the thin-film equation is challenging, even in the deterministic case;
including stochastic terms adds a further level of complexity.
To this aim, several sophisticated numerical methods have been developed, based on, e.g.,   
finite differences~\cite{diezGlobalModelsMoving2000}, finite elements~\cite{grunThinFilmFlowInfluenced2006} and spectral schemes~\cite{duran-olivenciaInstabilityRuptureFluctuations2019} .

@@

To simulate the dynamics of thin liquid films described by Eq.~(\ref{eq:thin_film}), we employ a recently developed~\cite{PhysRevE.100.033313} lattice Boltzmann method (LBM), built on 

@@

where $i$ labels the lattice velocities $c^{(i)}$ and runs from $0$ to $Q-1$, with $Q$ being the number of velocities characterizing the scheme, and $F$ is a (generalised) force\footnote{Strictly speaking its dimensions are 
$[\text{length}]^2[\text{time}]^{-2}$}.

@@

%c^{(i)}  = [0, 1, -1], \quad i = 0, 1, 2,

@@

with $g$ being the gravitational acceleration (which will be 
set to zero, hereafter, as it can be obviously neglected for nanoscale thin films).
The hydrodynamic fields (film height $h$ and velocity $u$) are expressed in terms of the 
distribution functions $f_i$ as:

@@

In the first term the effect of the film pressure $p$, Eq.~(\ref{eq:pressure}), is 
embedded, i.e. 

@@

depends on the film height and on the particular fluid velocity boundary 
condition at the substrate surface, parameterised by a slip length $\delta$.
It can be shown that in the limit of vanishing inertia, negligible gravity and small film thickness, the model defined by Eqs.~(\ref{eq:LBE_two}-\ref{eq:fric_alpha})
represents a numerical solver for the system of equations

@@

The left hand side of the second of Eqs.~(\ref{eq:lubr2eq}) can be neglected in the limits considered (small Reynolds number and film thickness). Thus, 
the equation reduces to $u \approx -\frac{h}{\mu\alpha_{\delta}(h)}\partial_x p$ that, plugged into the first of Eqs.~(\ref{eq:lubr2eq}), gives Eq.~(\ref{eq:thin_film}) 
(with a no-slip mobility $M(h)=\frac{h^3}{3\mu}$ if the further limit $\delta \rightarrow 0$ is taken).

@@

\subsection{Fluid-substrate interactions: the contact angle}\label{subsec:fluid_substrate}
For the disjoining pressure $\Pi(h)$ we use the following 
expression\textcolor{black}{\cite{crasterDynamicsStabilityThin2009, peschkaSignaturesSlipDewetting2019, nesicFullyNonlinearDynamics2015, oronLongscaleEvolutionThin1997}}:
\begin{equation}\label{eq:disjoin_p}
    \Pi(h) = \kappa(\theta) \left[\left(\frac{h_{\ast}}{h+h_c}\right)^n - \left(\frac{h_{\ast}}{h+h_c}\right)^m\right],
\end{equation}
where the prefactor $\kappa(\theta) = \gamma(1-\cos(\theta))\frac{(n-1)(m-1)}{(n-m)h_{\ast}}$ is related to the Hamaker constant by
$\mathcal{A} = 6 \pi h_{\ast}^3 \kappa(\theta)$.
$\theta$ is the contact angle, which effectively encodes the interfacial energies of the three-phase system and represents, therefore, a measure of the hydrophilicity/hydrophobicity of the substrate.
\textcolor{black}{Throughout the paper we restrict ourselves to low contact angles, $\theta < 1$,
so to comply with the lubrication approximation, which requires the derivative of the height field to be small, $\partial_x h(x) \ll 1$, and hence limits
the values of $\theta$ (since, at the contact point, $\partial_x h \sim \tan(\theta) \sim \theta$).}
\textcolor{black}{To ensure numerical stability we introduce a regularizing term $h_c$; the value $h_{\ast} - h_c > 0$, 
at which the disjoining pressure vanishes, sets the thickness of the precursor film covering the substrate.}

\section{Results}\label{sec:results_two}
\begin{center}
\begin{table}
    \centering
    \begin{tabular}{||l | c | c | p{1cm} | c | c | p{1cm} ||}
        \hline
        & \multicolumn{3}{ | c | }{$n=3, m=2$} & \multicolumn{3}{ | c | }{$n=9, m=3$} \\
        \hline
         $\theta$ & $t_0$ & $q_0$ & $\lambda_{\text{max}}$ & $t_0$ & $q_0$ & $\lambda_{\text{max}}$ \\ [0.5ex] 
         \hline\hline
         $\pi/9$ & $5 \cdot 10^5$ & $0.101$ & $62$ & $8.6 \cdot 10^6$ & $0.049$ & $128$ \\ 
         \hline
         $5\pi/36$ & $2 \cdot 10^5$ & $0.126$ & $50$ & $3.5 \cdot 10^6$ & $0.061$ & $103$  \\
         \hline
         $\pi/6$ & $10^5$ & $0.151$ & $42$ & $1.7 \cdot 10^6$ & $0.073$ & $86$ \\
         \hline
         $7\pi/36$ & $5 \cdot 10^4$ & $0.175$ & $36$ & $10^6$ & $0.085$ & $74$ \\
         \hline
         $2\pi/9$ & $3 \cdot 10^4$ & $0.2$ & $32$ & $6 \cdot 10^5$ & $0.097$ & $65$ \\ [1ex] 
         \hline
        \end{tabular}
        \caption{Numerical values (in lbu) of characteristic time, $t_0$, most unstable wavenumber, $q_0$, and corresponding wavelength, $\lambda_{\text{max}}$, 
        for the various contact angles $\theta$ and pairs $(n,m)$ of the disjoining pressure exponents used (see Eq.~(\ref{eq:disjoin_p}).}
    \label{tab:t_0s}
\end{table}
\end{center}
\textcolor{black}{We simulate numerically the model 
(\ref{eq:LBE_two}-\ref{eq:hydrofields}), with forcing term (\ref{eq:capillary_force}-\ref{eq:fric_force}-\ref{eq:thermal_force}-\ref{eq:tot_force}) (and $\Delta x = \Delta t = 1$), on a domain 
of length $L=4096$,
with a relaxation time $\tau = 1$, corresponding to a kinematic viscosity $\nu = 1/6$.
The surface tension $\gamma$ is set to $0.01$ and we use, as exponents appearing in the disjoining pressure (Eq. (\ref{eq:disjoin_p})),
the pairs $(n,m)=(3,2)$ (in section \ref{subsec:validation}, \ref{subsec:morphandrup} and \ref{subsec:results_contact_angle}) 
and $(n,m)=(9,3)$ (in section \ref{subsec:patterned_substrates}).
The thickness of the precursor film is $h_{\ast}-h_c = 0.05$ ($h_{\ast}=0.1, h_c = 0.05$) and the numerical slip length $\delta$ (unless otherwise stated)
is chosen to be in the weak slip regime, $\delta/h_0 = 1.0$. % The precursor film is only 0.05 in height, as there is as well hc added to h to make up 0.1
The system is initialised with zero velocity everywhere ($u(x,0)=0 \; \forall x \in [0, L]$)
and with random perturbations of amplitude $10^{-3}$ around the value $h_0 = 1$ for the film height, i.e. 
$h(x,0) = h_0 + 10^{-3} \varepsilon$ ($\varepsilon$ being a random variable uniformly distributed 
in $[-1, +1]$).
In the remainder of the manuscript all lengths will be given adimensionalised by $h_0$.
We vary the contact angle in the interval $[\pi/9, 2\pi/9 ]$;
the numerical values (in lattice Boltzmann units, lbu) of the corresponding wavelength, $\lambda_{\text{max}}$ and wavenumber, $q_0$, of the most unstable mode as well as the characteristic time, $t_0$, can be found in Tab.~\ref{tab:t_0s}.
For each contact angle configuration we perform one simulation for the athermal system and a set 
of $O(30)$ runs for the fluctuating case, at changing the random seed (the data shown represent, then, an average over the noise realisations). We would like to stress, incidentally, that the inclusion of 
the stochastic term entails a computational overhead of, roughly, $20\%$ of a typical deterministic run.\\ 
The numerical data will be indicated, hereafter, in terms of the dimensionless temperature $\sigma = \sqrt{\frac{k_B T}{\gamma h_0^2}}$, which is set either to $\sigma=0$ (for the athermal dewetting) or to $\sigma = 3.16 \cdot 10^{-3} \equiv \sigma_0$, 
corresponding to a thermal energy
of $k_B T = 10^{-7}$ lbu (for the fluctuating dewetting). For the sake of comparison, we remark 
that such a value lies well within the window $\sigma \in [10^{-3}, 10^{-2}]$, measured in experiments with polymeric and metallic thin films~\cite{fetzerThermalNoiseInfluences2007,doi:10.1021/la4009784}}.

>>>>

Wetting phenomena have an impact on areas as diverse as surface coatings, printing, plant treatment with pesticides, or even pandemics~\cite{oronLongscaleEvolutionThin1997, snoeijerMovingContactLines2013, cassieWettabilityPorousSurfaces1944, lenormandLiquidsPorousMedia1990,deryckGravityInertiaEffects1998, quereFluidCoatingFiber1999,bergeronControllingDropletDeposition2000,bhardwajLikelihoodSurvivalCoronavirus2020}.

@@

If this affinity is not sufficiently strong or the substrate is rather heterogeneous the liquid layer may eventually become unstable, leading to film rupture~\cite{oronLongscaleEvolutionThin1997, crasterDynamicsStabilityThin2009} or wrinkling~\cite{dasilvasobrinhoStudyDefectsUltrathin1999}. 

@@

A successful mathematical modelling of the dynamics of thin liquid films is based on the thin-film equation~\cite{reynoldsIIIExperimentalInvestigation1883, oronLongscaleEvolutionThin1997}

@@

The function $M(h)$ denotes the mobility, whose functional form is determined by the fluid velocity boundary condition at the solid surface: for a no-slip condition~\cite{oronLongscaleEvolutionThin1997}, in particular, $M(h)=\frac{h^3}{3\mu}$, where $\mu$ is the fluid dynamic viscosity.

@@

where $\gamma$ is the surface tension. The disjoining pressure $\Pi(h)$ is the derivative, with respect to the height $h$, of the interfacial potential, which incorporates the interactions between liquid and substrate (i.e. wetting properties).
The advantage of such approach in comparison with molecule-resolved methods such as Molecular Dynamics~\cite{haileMolecularDynamicsSimulation1997, zhangMolecularSimulationThin2019, wengMolecularDynamicsInvestigation2000, grabowThinFilmGrowth1988}
and Density Functional Theory~\cite{vanswolWettingDryingTransitions1989, tarazonaSimpleDensityFunctional1984, meisterDensityfunctionalTheoryInhomogeneous1985, hughesIntroductionInhomogeneousLiquids2014} is that these may quickly become computationally prohibitive, as the size of the film is increased above few nanometres in thickness and to the micrometric scale in the horizontal extension.

@@

For instance, in the context of dewetting, rupture times measured in experiments are shorter than predicted by simulations of the thin-film equation~\cite{bischofDewettingModesThin1996,herminghausSpinodalDewettingLiquid1998,beckerComplexDewettingScenarios2003}.

@@

Almost half a century later, a stochastic version of Eq.~(\ref{eq:thin_film}) was derived~\cite{grunThinFilmFlowInfluenced2006, meckeThermalFluctuationsThin2005, davidovitchSpreadingViscousFluid2005}, through a lubrication approximation of the Navier-Stokes-Landau-Lifshitz equations of fluctuating hydrodynamics~\cite{landauFluidMechanicsLandau2013}.

@@

The influence of thermal fluctuations on the short and long time morphology of a film has been studied with simulations of the one and two dimensional stochastic thin-film equations~\cite{nesicFullyNonlinearDynamics2015, pahlavanThinFilmsPartial2018}. 
Recently, new theoretical insights have been gained explaining film rupture as a consequence of the combined action of thermal fluctuations and drainage, both numerically~\cite{shahThermalFluctuationsCapillary2019} and experimentally~\cite{chatzigiannakisBreakupThinLiquid2020}.

@@

\section{Linear Stability Analysis of the Stochastic Thin-Film Equation}\label{sec:stoch_thin_film}
The one-dimensional stochastic thin-film equation reads~\cite{grunThinFilmFlowInfluenced2006, meckeThermalFluctuationsThin2005}:
\begin{equation}\label{eq:stoch_thin_film_simple}
    \partial_t h = \partial_x \left[\frac{h^3}{3\mu}\partial_x p + \sqrt{\frac{2k_BTh^3}{3\mu}}\mathcal{N}\right],
\end{equation}
where $k_B$ is Boltzmann's constant, $T$ is the temperature and $\mathcal{N}$ is a Gaussian white noise with
\begin{align}\label{eq:correlation}
    \langle\mathcal{N}(x,t)\rangle =&~0,\\ 
    \quad \langle\mathcal{N}(x,t)\mathcal{N}(x',t')\rangle =&~\delta(x-x')\delta(t-t').
\end{align}
In order to understand the influence of thermal fluctuations on the temporal evolution of the film height, we cursory recall the linear stability analysis of Eq.~(\ref{eq:stoch_thin_film_simple}) 
\cite{zhangMolecularSimulationThin2019, diezMetallicthinfilmInstabilitySpatially2016}.
Introducing the deviation $\delta h$ from the mean height $h_0$, such that $h = h_0 + \delta h$ and $\delta h \ll h_0$, 
the linearised stochastic thin-film equation is obtained
\begin{equation}\label{eq:linearstability_realspace}
    \partial_t \delta h = \frac{h_0^3}{3\mu}\Big(\Pi^{\prime}(h_0)\partial_x^2\delta h - \gamma \partial_x^4\delta h\Big) + \sqrt{\frac{2k_BT h_0^3}{3\mu}} \partial_x \mathcal{N},
\end{equation}
where $\Pi^{\prime}(h_0) \equiv \frac{\partial \Pi(h)}{\partial h}\bigg\rvert_{h=h_0}$.
Here, it is assumed that also the noise is small, in the sense that $k_B T \ll \gamma h_0^2$.
To derive a dispersion relation for this system, first the Fourier transforms of the perturbation $\delta h$  
\begin{equation}\label{eq:fourier_delta_h}
    \tilde{\delta h}(q,t) = \int_{-\infty}^{\infty} \delta h(x,t) e^{-ixq} dx,
\end{equation}
and the noise term $\mathcal{N}$
\begin{equation}\label{eq:fourier_noise}
    \tilde{\mathcal{N}}(q,t) = \int_{-\infty}^{\infty} \mathcal{N}(x,t) e^{-ixq} dx.
\end{equation}
must be performed.

Inserting $\tilde{\delta h}$ and $\tilde{\mathcal{N}}$ in~(\ref{eq:linearstability_realspace}) yields
\begin{equation}\label{eq:stability_fourier}
    \partial_t \tilde{\delta h} = \omega(q)\tilde{\delta h} +i\sqrt{\frac{2k_BTh_0^3}{3\mu}}q\tilde{\mathcal{N}},
\end{equation}
with dispersion relation $\omega(q)$ given by
\begin{equation}\label{eq:dispersion}
    \omega(q) = \frac{1}{t_0}\left[2\left(\frac{q}{q_0}\right)^2 - \left(\frac{q}{q_0}\right)^4\right].
\end{equation}
The maximum, $q_0$, of $\omega(q)$ and the characteristic time, $t_0$, are specific to the substrate the film is deposited on and their expressions are~\cite{fetzerThermalNoiseInfluences2007}
\begin{equation}\label{eq:q0}
    q_0^2 = \frac{1}{2\gamma}\frac{\partial \Pi(h)}{\partial h}\bigg\rvert_{h=h_0},
\end{equation}
and
\begin{equation}\label{eq:t0}
    t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}.
\end{equation}
Solving Eq.~(\ref{eq:stability_fourier}) leads to the following expression for the spectrum~\cite{zhangMolecularSimulationThin2019,meckeThermalFluctuationsThin2005},
$S(q,t) = \langle |\tilde{\delta h}(q,t)|^2\rangle$ (where the brackets stand for the average over the noise):
\begin{equation}\label{eq:structure_factor}
    S(q,t) = S_0(q)e^{2\omega(q)t} + \frac{k_BTh_0^3}{3\mu}\frac{q^2}{\omega(q)}(e^{2\omega(q)t} - 1).
\end{equation}
For $q>\sqrt{2}q_0$, $S(q,t)$ tends to the capillary wave spectrum, 
$S \sim S_{\text{cw}} \propto \frac{k_BT}{\gamma q^2}$, as $t \rightarrow \infty$ (whereas it would decay exponentially to zero in the deterministic case)~\cite{fetzerThermalNoiseInfluences2007,meckeThermalFluctuationsThin2005}.
Moreover, while the maximum $q_0$ is independent of time in the deterministic case, the maximum $q_m$ of~(\ref{eq:structure_factor}) approaches $q_0$ from the right as the time increases.
It follows that the most unstable wavelength, $\lambda_{\text{max}} = 2\pi/q_m$, grows with time.
We will come back to this fact in Sec.~\ref{sec:results_two} and show that indeed this behavior is reproduced and observed in our simulations.

\section{Thin Film Lattice Boltzmann Model}\label{sec:num_method}
Performing numerical simulations of the thin-film equation is challenging, even in the deterministic case; including stochastic terms adds a further level of complexity.
To this aim, several sophisticated numerical methods have been developed, based on, e.g., finite differences~\cite{diezGlobalModelsMoving2000}, finite elements~\cite{grunThinFilmFlowInfluenced2006} and spectral schemes~\cite{duran-olivenciaInstabilityRuptureFluctuations2019}.

@@

To simulate the dynamics of thin liquid films described by Eq.~(\ref{eq:thin_film}), we employ a recently developed~\cite{zitzLatticeBoltzmannMethod2019} lattice Boltzmann method (LBM), built on 

@@

where $i$ labels the lattice velocities $c^{(i)}$ and runs from $0$ to $Q-1$, with $Q$ being the number of velocities characterizing the scheme, and $F$ is a (generalised) force\footnote{Strictly speaking its dimensions are $[\text{length}]^2[\text{time}]^{-2}$}.

@@

@@

with $g$ being the gravitational acceleration (which will be set to zero, hereafter, as it can be obviously neglected for nanoscale thin films).
The hydrodynamic fields (film height $h$ and velocity $u$) are expressed in terms of the distribution functions $f_i$ as:

@@

In the first term the effect of the film pressure $p$, Eq.~(\ref{eq:pressure}), is embedded, i.e. 

@@

depends on the film height and on the particular fluid velocity boundary condition at the substrate surface, parameterised by a slip length $\delta$.
It can be shown that in the limit of vanishing inertia, negligible gravity and small film thickness, the model defined by Eqs.~(\ref{eq:LBE_two}-\ref{eq:fric_alpha}) represents a numerical solver for the system of equations

@@

The left hand side of the second of Eqs.~(\ref{eq:lubr2eq}) can be neglected in the limits considered (small Reynolds number and film thickness). 
Thus, the equation reduces to $u \approx -\frac{h}{\mu\alpha_{\delta}(h)}\partial_x p$ that, plugged into the first of Eqs.~(\ref{eq:lubr2eq}), gives Eq.~(\ref{eq:thin_film}) (with a no-slip mobility $M(h)=\frac{h^3}{3\mu}$ if the further limit $\delta \rightarrow 0$ is taken).

@@

\subsection{Fluid-substrate interactions: the contact angle}\label{subsec:fluid_substrate}
For the disjoining pressure $\Pi(h)$ we use the following expression~\cite{crasterDynamicsStabilityThin2009, peschkaSignaturesSlipDewetting2019, nesicFullyNonlinearDynamics2015, oronLongscaleEvolutionThin1997}:
\begin{equation}\label{eq:disjoin_p}
    \Pi(h) = \kappa(\theta) \left[\left(\frac{h_{\ast}}{h+h_c}\right)^n - \left(\frac{h_{\ast}}{h+h_c}\right)^m\right],
\end{equation}
where the prefactor $\kappa(\theta) = \gamma(1-\cos(\theta))\frac{(n-1)(m-1)}{(n-m)h_{\ast}}$ is related to the Hamaker constant by $\mathcal{A} = 6 \pi h_{\ast}^3 \kappa(\theta)$.
$\theta$ is the contact angle, which effectively encodes the interfacial energies of the three-phase system and represents, therefore, a measure of the hydrophilicity/hydrophobicity of the substrate.
Throughout the paper we restrict ourselves to low contact angles, $\theta < 1$, so to comply with the lubrication approximation, which requires the derivative of the height field to be small, $\partial_x h(x) \ll 1$, and hence limits the values of $\theta$ (since, at the contact point, $\partial_x h \sim \tan(\theta) \sim \theta$).
To ensure numerical stability we introduce a regularizing term $h_c$; the value $h_{\ast} - h_c > 0$, at which the disjoining pressure vanishes, sets the thickness of the precursor film covering the substrate.

\section{Results}\label{sec:results_two}
\begin{center}
\begin{table}
    \centering
    \begin{tabular}{||l | c | c | p{1cm} | c | c | p{1cm} ||}
        \hline
        & \multicolumn{3}{ | c | }{$n=3, m=2$} & \multicolumn{3}{ | c | }{$n=9, m=3$} \\
        \hline
         $\theta$ & $t_0$ & $q_0$ & $\lambda_{\text{max}}$ & $t_0$ & $q_0$ & $\lambda_{\text{max}}$ \\ [0.5ex] 
         \hline\hline
         $\pi/9$ & $5 \cdot 10^5$ & $0.101$ & $62$ & $8.6 \cdot 10^6$ & $0.049$ & $128$ \\ 
         \hline
         $5\pi/36$ & $2 \cdot 10^5$ & $0.126$ & $50$ & $3.5 \cdot 10^6$ & $0.061$ & $103$  \\
         \hline
         $\pi/6$ & $10^5$ & $0.151$ & $42$ & $1.7 \cdot 10^6$ & $0.073$ & $86$ \\
         \hline
         $7\pi/36$ & $5 \cdot 10^4$ & $0.175$ & $36$ & $10^6$ & $0.085$ & $74$ \\
         \hline
         $2\pi/9$ & $3 \cdot 10^4$ & $0.2$ & $32$ & $6 \cdot 10^5$ & $0.097$ & $65$ \\ [1ex] 
         \hline
        \end{tabular}
        \caption{Numerical values (in lbu) of characteristic time, $t_0$, most unstable wavenumber, $q_0$, and corresponding wavelength, $\lambda_{\text{max}}$, 
        for the various contact angles $\theta$ and pairs $(n,m)$ of the disjoining pressure exponents used (see Eq.~(\ref{eq:disjoin_p}).}
    \label{tab:t_0s}
\end{table}
\end{center}
We simulate numerically the model (\ref{eq:LBE_two}-\ref{eq:hydrofields}), with forcing term (\ref{eq:capillary_force}-\ref{eq:fric_force}-\ref{eq:thermal_force}-\ref{eq:tot_force}) (and $\Delta x = \Delta t = 1$), on a domain of length $L=4096$, with a relaxation time $\tau = 1$, corresponding to a kinematic viscosity $\nu = 1/6$.
The surface tension $\gamma$ is set to $0.01$ and we use, as exponents appearing in the disjoining pressure (Eq. (\ref{eq:disjoin_p})), the pairs $(n,m)=(3,2)$ (in section \ref{subsec:validation}, \ref{subsec:morphandrup} and \ref{subsec:results_contact_angle}) and $(n,m)=(9,3)$ (in section \ref{subsec:patterned_substrates}).
The thickness of the precursor film is $h_{\ast}-h_c = 0.05$ ($h_{\ast}=0.1, h_c = 0.05$) and the numerical slip length $\delta$ (unless otherwise stated) is chosen to be in the weak slip regime, $\delta/h_0 = 1.0$. 
The system is initialised with zero velocity everywhere ($u(x,0)=0 \; \forall x \in [0, L]$) and with random perturbations of amplitude $10^{-3}$ around the value $h_0 = 1$ for the film height, i.e. $h(x,0) = h_0 + 10^{-3} \varepsilon$ ($\varepsilon$ being a random variable uniformly distributed in $[-1, +1]$).
In the remainder of the manuscript all lengths will be given adimensionalised by $h_0$.
We vary the contact angle in the interval $[\pi/9, 2\pi/9 ]$; the numerical values (in lattice Boltzmann units, lbu) of the corresponding wavelength, $\lambda_{\text{max}}$ and wavenumber, $q_0$, of the most unstable mode as well as the characteristic time, $t_0$, can be found in Tab.~\ref{tab:t_0s}.
For each contact angle configuration we perform one simulation for the athermal system and a set of $O(30)$ runs for the fluctuating case, at changing the random seed (the data shown represent, then, an average over the noise realisations). 
We would like to stress, incidentally, that the inclusion of the stochastic term entails a computational overhead of, roughly, $20\%$ of a typical deterministic run.\\ 
The numerical data will be indicated, hereafter, in terms of the dimensionless temperature $\sigma = \sqrt{\frac{k_B T}{\gamma h_0^2}}$, which is set either to $\sigma=0$ (for the athermal dewetting) or to $\sigma = 3.16 \cdot 10^{-3} \equiv \sigma_0$, corresponding to a thermal energy
of $k_B T = 10^{-7}$ lbu (for the fluctuating dewetting). 
For the sake of comparison, we remark that such a value lies well within the window $\sigma \in [10^{-3}, 10^{-2}]$, measured in experiments with polymeric and metallic thin films~\cite{fetzerThermalNoiseInfluences2007,gonzalezInstabilityLiquidCu2013}.
